Tutkielman otsikko

### 2020-02-07 gradu/msc_english_contents.tex
adds more text

Variable-byte encoding \citep{Wil99} (VB) is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
and allows constant time random access (TODO: link to later chapters). A good data set for VB encoding is a list of mostly small numbers with a 
need to support larger ones.

A search engine may use an inverted index of words in documents. For each word, a list of document IDs where the word appears is stored. 
It may also store locations of the word in document for advanced search purposes. Usually these lists are preprocessed and stored 
as gaps, storing the difference to previous number instead of the actual number \citep{Man08}. Common words have a lot of entries in these lists, 
but because of gap storing the numbers are small. In contrast, rare words have only a few entries but the numbers stored are larger. These lists 
are excellent data sets for VB. Apache Lucene has variable bytes as vInt and IBM DB2 has them as Variable Byte).

TODO: a chapter about existing work
Usage of VB to compress inverted 
index lists was first experimented in \citep{Sch02}. (TODO: find better examples or find references to these)

VB splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.

@@

+description

\medskip
-motivation (see papers below)

\medskip
+pseudocode

\medskip
+a few examples (1)

\medskip
-a review of related work should include:

\medskip
---the text book from which the initial description I sent you comes from

\medskip
---First use for inverted index compression (at least that I'm aware of):

\medskip
https://dl.acm.org/doi/10.1145/564376.564416

\medskip
---More recent incarnations of vbyte (just mention them):

\medskip
https://arxiv.org/abs/1709.08990

\medskip
https://arxiv.org/abs/1503.07387

\medskip

>>>>

Variable-byte (VB) encoding \citep{Wil99} is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
and allows constant time random access. This is later explained in (TODO: link to later chapter).

A good data set for VB encoding is a list of mostly small numbers with a need to support larger ones. A search engine may use an inverted index of 
words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
advanced search purposes. Usually these lists are preprocessed and stored as gaps, storing the difference to previous number instead of the actual 
number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. In contrast, rare words 
have only a few entries but the numbers stored are larger. These lists are excellent data sets for VB encoding.

VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.

@@

VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar data types have existed for example in Apache Lucene 
(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was found efficient in compressing integer lists. It was first used to compress
lists of word locations in documents \citep{Sch02}. It yielded excellent records, and since then many different approaches have been introduced.

history and current implementations etc here
 - mainitse IR book, Lem15, Pla18

TODO: a chapter about existing work
